---
output: github_document
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# namedness

```{r}
# load all internals
devtools::load_all()
```


Exploring the `MAYBE_REFERENCED()` and `MAYBE_SHARED()` interaction with Rcpp.

Currently on R 3.5.3.

## Examples

If you just use `check_namedness()` as Rcpp exports it, we have a problem where, no matter what, it looks like the objects are referenced.

This correctly says that `x` has a name attached to it.

```{r}
x <- c(1, 2, 3)
check_namedness(x)
```

But when we move to temporary (non ALTREP) objects, these shouldn't be named:

```{r}
check_namedness(c(1, 2, 3))
```

The problem is the fact that we have an R function wrapping the `.Call()`. If we use the `.Call()` mechanism directly, all seems well.

```{r}
.Call(`_namedness_check_namedness`, c(1, 2, 3))
```

If we use an ALTREP-like object, it will be referenced, by design.

```{r}
.Call(`_namedness_check_namedness`, 1)
.Call(`_namedness_check_namedness`, 1L)
.Call(`_namedness_check_namedness`, 1:2)

# Look at NAM(3)!
.Internal(inspect(1L))

# NAM(3) and (compact)
.Internal(inspect(1:2))

# As opposed to this:
# No NAM() tag
.Internal(inspect(c(1, 2, 3)))
```

What about temporaries?

```{r}
x <- c(1, 2, 3)
y <- c(3, 4, 5)

# of course...
check_namedness(x + y)

# nothing here!
.Call(`_namedness_check_namedness`, x + y)
```

This idea is useful because we could mimic base R (in theory), and reuse the memory of temporaries when performing operations like `x + y + z`. Reuse the memory of `x + y = <temporary>` to compute `<temporary> + z` (as long as the result has the right shape).


But, we obviously need to wrap this inside a function...which is basically what Rcpp does for us. And we immediately hit the issue again.

```{r}
nmdness <- function(x) .Call(`_namedness_check_namedness`, x)
nmdness(c(1, 2))
```

There is a note here on the fact that `NAMED = 1` optimization only really works effectively if the function is a primitive. "Note that optimizing NAMED = 1 is only effective within a primitive (as the closure wrapper of a .Internal will set NAMED = NAMEDMAX when the promise to the argument is evaluated) and hence replacement functions should where possible be primitive to avoid copying (at least in their default methods)."
https://cran.r-project.org/doc/manuals/r-release/R-ints.html#g_t_002eInternal-vs-_002ePrimitive

Another note is in R Extensions, in 5.9.10 here at Named Objects and Copying. "The case NAMED(foo) == 1 allows some optimization, but it can be ignored (and duplication done whenever NAMED(foo) > 0). (This optimization is not currently usable in user code.) It is intended for use within replacement functions." 
https://cran.r-project.org/doc/manuals/r-release/R-exts.html#Named-objects-and-copying

This still doesn't suggest how one could use the `NAMED = 0` case though.

I think this also applies to the `NAMED = 0` case with temporaries. If we wrap it in a function, we immediately have issues.

I guess `NAMED = 0` could be useful if you pass a SEXP created at the C++ level to another C++ function immediately, which could take advantage of the fact that you haven't bound it to any name at the R level. Perhaps this is what `vctrs::vec_restore()` is doing when being called from something like `vec_slice()`.
https://github.com/r-lib/vctrs/blob/8251896d6db132ada6e7702b753de24dee66dc56/src/slice.c#L142
https://github.com/r-lib/vctrs/blob/84839833c78b5df23a769951cff62f37d9d43dec/src/cast.c#L278

Note that all the base operators like `+` are primitives, so they can take advantage of `NO_REFERENCES()` and reuse memory smartly.

## Solution

I don't think there is one right now. Obviously we can't have users using `.Call()`, but it seems impossible to wrap the `.Call()` in something more user friendly to use where we can still take advantage of having "no references".

Here is another example from Dirk showing the same issue:
https://github.com/RcppCore/Rcpp/issues/943#issuecomment-469231746


